---
title: "bike_sharing_report"
author: "Mila"
date: "2022-08-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Install and load the necessary packages.
```{r intro}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
```
Check your current directory.
```{r directory}
getwd()
```
Change your working directory if necessary with setwd("new_directory").

## Data Collection
I have previously cleaned and combined data from 12 months into four quarters. Let's combine now that data into one year. 
```{r data_merging}
q1 <- read_csv("q1.csv")
q2 <- read_csv("q2.csv")
q3 <- read_csv("q3.csv")
q4 <- read_csv("q4.csv")
yearly_trips <- bind_rows(q1,q2,q3,q4)
```
Check the column names and data types with 
```{r data_type_check}
sapply(yearly_trips, typeof)
```
I have previously deleted unnecessary rows with
```{r eval=FALSE}
yearly_trips <- yearly_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))

```
Add the columns that will list date, month, day of the week, and year of the ride. This will make data aggregation easier.
```{r adding_columns}
yearly_trips$date <- as.Date(yearly_trips$started_at)
yearly_trips$month <- format(as.Date(yearly_trips$date), "%m")
yearly_trips$day <- format(as.Date(yearly_trips$date), "%d")
yearly_trips$year <- format(as.Date(yearly_trips$date), "%Y")
yearly_trips$day_of_week <- format(as.Date(yearly_trips$date), "%A")
```
Add a ride_length column to the data.

```{r warning=FALSE}
yearly_trips$ride_length <- difftime(yearly_trips$ended_at,yearly_trips$started_at)
```

There are going to be rows with negative ride lengths. Delete them with

```{r}
yearly_trips <- yearly_trips[!(yearly_trips$ride_length<=0),]

```

You can also get rid of some rows containing blank or N/A values with
```{r eval=FALSE}
yearly_trips <- yearly_trips[!(is.na(yearly_trips$start_station_id) | yearly_trips$start_station_id==""), ]
```

## Aggregation and Visualization
Make a new dataset that aggreates weekday and membership type.
```{r message=FALSE, warning=FALSE}
trips_weekday <- yearly_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)

```
Visualize the number of rides by rider type depending on weekday.

```{r}
trips_weekday %>% ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

Do the same but with average duration for trips.

```{r message=FALSE, warning=FALSE}
trips_weekday %>% ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

Now, let's perform another data aggregation but this time use month as a parameter instead of weekday.

```{r message=FALSE, warning=FALSE}
trips_month <- yearly_trips %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month)

```
Make plots for this new dataset as well.
```{r}
trips_month %>% ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
```{r}
trips_month %>% ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

All done! Save the plots and proceed to share your data to gain more insigts.


